---
title: |
  | Statistical analysis of data from an 
  | anthocyanin measurement experiment in 
  | *A. thaliana* seedlings
author: "CrisprCat"
output: 
  pdf_document: 
    toc: true
---
# Introduction
This script describes the statistical analysis of an experiment in which anthocyanin levels in 5 day old *A. thaliana* seedlings grown in red light were determined. The anthocyanin levels of 3 different mutants and wildtype seedlings were analysed. To identify statistical significant differences among the means of anthocyanin levels of the mutants an One-Way ANOVA is used.

The data used for this example is available under DOI: 10.1038/s41477-020-0725-0 and is the source data of Figure 1d.

# Setup
## Installing and loading required packages

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# Store package names, required for the analysis in a vector
packages <- c("tidyverse", "car", "multcomp")

# Install packages that are not yet installed
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Load packages
invisible(lapply(packages, library, character.only = TRUE))
```  

# Load the data
The data is usually stored in a .csv file with one column describing the "Gentoype" and a second column describing the measured value "anthocyanin_level".

```{r}
# Read the .csv file
antho <- read.csv(file = "anthocyanin_levels.CSV", sep = ";", header = T)

# Change the class of Genotype from character to factor
antho$Genotype <- as.factor(antho$Genotype)

# Transform the data
#   The log transformation is required so that the data meets the assumptions for 
#   an ANOVA
antho$T_log <- log(antho$anthocyanin_level) 
                                            
```

# Statistical analysis
Key assumptions for the One-Way ANOVA are, that the data  
* follows a normal distribution  
* shows homogeneity of variances

As this biological data is independently and randomly sampled from a population I assume that these requirements are met after log transformation. However, it is also possible to test for these assumptions. 

## Test the assumption of equal variances
With the Brown-Forsythe test the null hypothesis, that the variances of the analysed groups (anthocyanin levels in each genotype) are equal, is tested.

### Brown-Forsythe test
The Brown-Forsythe test is more robust to not gaussian distributed data than the Levene's test.
```{r}
# Test for homoscedasticity
leveneTest(antho$T_log ~ antho$Genotype, center = median)
```
As the p-value (Pr) is > 0.05, the null hypothesis cannot be rejected.

## Test the assumption of normal distribution
The assumption of normal distribution of the data in each group = (genotype) can be graphically tested with a QQ-plot or with the Shapiro-Wilk's test.

### QQ-plot
When the points in the QQ-plot show linearity it suggests the data is normally distributed.
```{r}
# Create a QQ-plot for WT anthocyanin levels
qqnorm(antho$T_log[antho$Genotype == 'WT'])
qqline(antho$T_log[antho$Genotype == 'WT'])

# Create a QQ-plot for hy5 anthocyanin levels
qqnorm(antho$T_log[antho$Genotype == 'hy5'])
qqline(antho$T_log[antho$Genotype == 'hy5'])

# Create a QQ-plot for bbx202122 anthocyanin levels
qqnorm(antho$T_log[antho$Genotype == 'bbx202122'])
qqline(antho$T_log[antho$Genotype == 'bbx202122'])

# Create a QQ-plot for bbx202122hy5 anthocyanin levels
qqnorm(antho$T_log[antho$Genotype == 'bbx202122hy5'])
qqline(antho$T_log[antho$Genotype == 'bbx202122hy5'])
```

### Shapiro-Wilk's test
With the Shapiro-Wilk's test the null hypothesis, that the the samples in each analysed group (anthocyanin levels in each genotype) follow a normal distribution, is tested.

```{r}
# Test for normal distribution
shapiro.test(antho$T_log[antho$Genotype == 'WT'])

# Test for normal distribution
shapiro.test(antho$T_log[antho$Genotype == 'hy5'])

# Test for normal distribution
shapiro.test(antho$T_log[antho$Genotype == 'bbx202122'])

# Test for normal distribution
shapiro.test(antho$T_log[antho$Genotype == 'bbx202122hy5'])
```
As the p-value (Pr) in each group is > 0.05, the null hypothesis cannot be rejected.

## One-Way ANOVA
With the One-Way ANOVA the null hypothesis, that there are no statistical significant differences among the means of the anthocyanin levels of the mutants, is tested.
  
```{r}
# Fit an Analysis of Variance model
res.aov <- aov(T_log ~ Genotype, data = antho)

# Show the results of the fitted model
summary(res.aov)
```
As the p-value (Pr) is < 0.05, the null hypothesis can be rejected.


## Multiple comparisons
To determine if the means of specific groups are statistically significant different from each other the Tukey HSD post-hoc test is computed.

```{r}
# Perform the post-hoc test
TukeyHSD(res.aov, conf.level = 0.95)
```
When the p adj value is < 0.05 the null hypothesis, that the means of the two compared groups are not statistically significant different, is rejected.

To assign letters to each group, that indicate groups that are statistically significant different from each other the cld() function is used.

```{r}
# Create a glht (general linear hypotheses) object
Tukey_results = glht(res.aov, linfct = mcp(Genotype = "Tukey"))

# Create a cld (compact letter display) object of the pairwise comparisons
tuk_cld <- cld(Tukey_results, decreasing = TRUE)

# Display the letters assigned to each group 
print(tuk_cld)
```

# Data visualization
## Calculate summary statistics
These summary statistics are used to create a meaningful data visualization of the experiments result. 

```{r}
# Calculate summary statistics
antho_summary = antho %>%
  group_by(Genotype) %>%
  summarise(mean_antho = mean(anthocyanin_level), # Calculate the mean of anthocyanin
                                                  # levels for each genotype
            sd_antho = sd(anthocyanin_level), # Calculate the standard deviation of 
                                              # anthocyanin levels for each genotype
            n_antho = n(), # Calculate the number of measurements of anthocyanin levels
                           # for each genotype
            SE_antho = sd_antho / sqrt(n()), # Calculate the standard error of the mean
                                             # of anthocyanin levels for each genotype
            max_antho = max(anthocyanin_level)) %>% # Define the maximum of anthocyanin
                                                    # levels for each genotype
  # Add a column with the letters indicating statistical significant different groups
  mutate(diff = print(tuk_cld$mcletters$Letters)) 

# Display the summary statistics
print(antho_summary)
```

## Create a plot and save it as a pdf

```{r}
# Save the graph as a pdf
pdf("anthocyanin_measurement.pdf", width = 4 , height = 6)
# Create a plot with the mean of the anthocyanin level vs. Genotype
antho_plot = ggplot(data = antho_summary, aes(x = Genotype, y = mean_antho)) +
  # Create a bar plot and customize its appearance
  geom_bar(stat = "identity", # Create the bars 
           fill = c("#FFFFFF", "#FF6161",
                    "#9B8EFF", "#787878"), # assign individual fill colors to the bars
           color = "black") + 
  geom_errorbar(aes(ymin = mean_antho - SE_antho, ymax = mean_antho + SE_antho),
                width = 0.2) + # add error bars
  # Change order of the categories
  scale_x_discrete(limits = c("WT", "hy5", "bbx202122", "bbx202122hy5")) + 
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90,
                                   size = 12,
                                   hjust = 1, 
                                   vjust = 0.5, 
                                   face = c("plain", "italic", "italic", "italic"), 
                                   color = "black"),
        axis.title.y = element_text(size = 12, 
                                    color = "black"),
        axis.text.y = element_text(size = 12, 
                                   color = "black")) +
  theme(axis.ticks = element_line()) +
  ylab("Anthocyanin (per g fresh weight)") +
  xlab(NULL) +
  scale_y_continuous(breaks = c(0, 1, 2, 3), 
                     expand = c(0,0), 
                     limits = c(0,3)) +
  geom_text(data = antho_summary, 
            aes(y = max_antho, label = diff), 
            vjust = -1, 
            size = 6, 
            color = "black") +
  geom_jitter(data = antho, 
              aes(y = anthocyanin_level), 
              size = 4, 
              color = "black", 
              shape = 1, 
              width = 0.2)


antho_plot
dev.off()

# Display the graph
print(antho_plot)

```